---
title: "Bayesian Final Project"
author: "Tianyi Li, Jade Gu, Jiashu Liu, Jeremy Lu"
date: "5/12/2023"
output: 
  pdf_document:
  latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs, include=TRUE, message=FALSE, warning=FALSE}
# Load packages
library(bayesrules)
library(tidyverse)
library(janitor)
library(ggplot2)
library(rstan)
library(bayesplot)
library(broom.mixed)
library(rstanarm)
library(tidybayes)
```

## 2.1 Proposal

In this project we will be developing bayesian models in order to identify the effects of mother's characteristics including her education level, race, age, gestational length, smoking status, and BMI, on infant's birth weight.

## 2.2 Attribution 

Tianyi Li: Project leader, team management, raw data summary, statistical model

Jade Gu: Raw data summary, model fitting and selection, assisting in prior predictive simulation

Jiashu Liu:  Prior predictive simulation, assisting in model fitting

Jeremy Lu: Report writing, presentation preparation, assist in summarizing key convergence diagnostics 


## 2.3 Raw Data Summary

### 2.3.1 Data Origin

Below is a link to a compressed CSV file containing US birth data for the year 2018. The data is provided by the National Center for Health Statistics and can be accessed from the National Bureau of Economic Research website. 

```{r loading, include=FALSE}
#url <- "https://data.nber.org/natality/2018/natl2018.csv.zip"
# get full 2018 birth dataset
#birth_full <- read.csv(url)
birth_full <- read.csv("~/Desktop/2123 Bayesian Inference/project/natl2018us.csv")
```

### 2.3.2 Data Clean (Include Variable Selection)

This 2018 birth data contains 3801534 obs. of 240 variables. We will pick 7 most interesting variables including: dbwt(infant's birthweight in grams), mager(mother's single years of age), mracehisp(Mother’s Race/Hispanic Origin), meduc(mother's education), cig_rec(cigarette recode), bmi_r(body mass index recode), combgest(combined gestation - detail in weeks). Then we will filter to exclude any unknown or not stated pieces.

```{r, include=FALSE}
set.seed(84735)

# Check all variables name
names(birth_full)
# Select 7 most interesting/key variables
birth_clean <- birth_full %>% 
  select(dbwt, mager, mracehisp, meduc, 
         cig_rec, bmi_r, combgest)
# keep only complete cases
birth_clean <- birth_clean %>% 
  filter(dbwt != 9999, 
         mracehisp != 8,
         meduc != 9, 
         cig_rec != 'U', 
         bmi_r != 9, 
         combgest != 99
         )
```

### 2.3.3 Data Sampling

Since this birth data after filter contains 3643397 obs. which is still extremely large, we will get a random sample of births that includes only 0.1% of all obs.

```{r, include=FALSE}
set.seed(84735)
# Get a random sample includes 0.1% of all observations in the whole dataset
birth_sample <- birth_clean[sample(1:nrow(birth_clean), nrow(birth_clean)*0.001),]

# save it as a csv file
# write.csv(birth_sample, file = "~/Desktop/2123 Bayesian Inference/project/birth.csv", row.names = FALSE)
```

### 2.3.4 Exploratory data analysis
```{r, include=FALSE}
#summary statistics for the numeric variables
```
**dbwt**
**mager**
**mracehisp**
**meduc**
**cig_rec**
**bmi_r**
**combgest**


### Explanatory data analysis (Some intersting features).
```{r}
birth <- birth_sample
```

```{r}
ggplot(birth, aes(x = combgest, y = dbwt)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
  labs(x = "combgest", y = "dbwt")
```
There is a positive linear relationship between mother's combined gestation time and infant's birthweight. 


```{r}
ggplot(birth, aes(x = combgest, y = dbwt, color = cig_rec)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```



## 2.4 Statistical Model

$Y_i$ is a continuous variable


```{r}
m1 <- stan_glm(
  dbwt ~ mager + mracehisp + meduc + cig_rec + bmi_r + combgest,
  data = birth, family = gaussian, 
  prior_intercept = normal(4000,250),
  prior = normal(200, 15, autoscale = TRUE), 
  prior_aux = exponential(1, autoscale = TRUE),
  chains = 4, iter = 5000*2, seed = 84735,
  prior_PD = FALSE)
```

```{r}
# Visual diagnostics of MCMC simulation
mcmc_trace(m1, size = 0.1)
mcmc_dens_overlay(m1)
mcmc_acf(m1)

# Numerical diagnostics of MCMC simulation
neff_ratio(m1)
rhat(m1)
```
