---
title: "Bayesian Final Project"
author: "Tianyi Li, Jade Gu, Jiashu Liu, Jeremy Lu"
date: "5/12/2023"
output: 
  pdf_document:
  latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs, include=TRUE, message=FALSE, warning=FALSE}
# Load packages
library(bayesrules)
library(tidyverse)
library(janitor)
library(ggplot2)
library(rstan)
library(bayesplot)
library(broom.mixed)
library(rstanarm)
library(tidybayes)
```

## 2.1 Proposal

In this project we will be developing bayesian models in order to identify the effects of mother's characteristics including her education level, race, age, gestational length, smoking status, and BMI, on baby girl's birth weight.

## 2.2 Attribution 

Tianyi Li: Project leader, team management, raw data summary, statistical model

Jade Gu: Raw data summary, model fitting and selection, assisting in prior predictive simulation

Jiashu Liu: Prior predictive simulation, assisting in model fitting

Jeremy Lu: Report writing, presentation preparation, assist in summarizing key convergence diagnostics 


## 2.3 Raw Data Summary

### 2.3.1 Data Origin

Below is a link to a compressed CSV file containing US birth data for the year 2018. The data is provided by the National Center for Health Statistics and can be accessed from the National Bureau of Economic Research website. 

```{r loading, include=FALSE}
#url <- "https://data.nber.org/natality/2018/natl2018.csv.zip"
# get full 2018 birth dataset
#birth_full <- read.csv(url)
birth_full <- read.csv("~/Desktop/2123 Bayesian Inference/project/natl2018us.csv")
```

### 2.3.2 Data Clean (Include Variable Selection)

This 2018 birth data contains 3801534 obs. of 240 variables. First, based on our research topic, we will keep girls' data only. Second, we will pick 8 most interesting variables including: dbwt(infant's birthweight in grams), mager(mother's single years of age), dmar(martial status), mracehisp(Mother’s Race/Hispanic Origin), meduc(mother's education), cig_rec(cigarette recode), bmi(body mass index), combgest(combined gestation - detail in weeks). Then, we will exclude any unknown or not stated piece. After that, we will factor all categorical variables.

```{r, include=FALSE}
set.seed(84735)

# Check all variables name
names(birth_full)

# Keep girls' newborns' data
birth_clean <- birth_full %>% 
  filter(sex != 'M')

# Select 8 most interesting/key variables
birth_clean <- birth_clean %>% 
  select(dbwt, mager, dmar, mracehisp, 
         meduc, cig_rec, bmi, combgest)

# Exclude any unknown or not stated pieces
birth_clean <- birth_clean %>% 
  filter(dbwt != 9999, 
         dmar != 9,
         mracehisp != 8,
         meduc != 9, 
         cig_rec != 'U', 
         bmi != 99.9, 
         combgest != 99
         )

# Factor categorical variables
birth_clean <- birth_clean %>% 
  mutate(dmar = as.factor(dmar),
         mracehisp = as.factor(mracehisp), 
         meduc = as.factor(meduc), 
         cig_rec = as.factor(cig_rec))
```

### 2.3.3 Data Sampling

Since this birth data after filter contains 1576269 obs. which is still extremely large, we will get a random sample of births that includes only 0.1% of all obs.

```{r, include=FALSE}
set.seed(84735)
# Get a random sample includes 0.1% of all observations in the whole dataset
birth_sample <- birth_clean[sample(1:nrow(birth_clean), nrow(birth_clean)*0.001),]

# save it as a csv file
write.csv(birth_sample, file = "~/Desktop/2123 Bayesian Inference/project/birth.csv", row.names = FALSE)
```

```{r, include=FALSE}
# load sampled data
birth <- read.csv("~/Desktop/2123 Bayesian Inference/project/birth.csv")
```

### 2.3.4 Exploratory data analysis
```{r, include=FALSE}
#summary statistics for the numeric variables
```
**dbwt**
**mager**
**dmar**
**mracehisp**
**meduc**
**cig_rec**
**bmi**
**combgest**


## 2.4 Statistical Model
## 2.5 Prior Predictive Simulation
## 2.6 Modeling Fitting, PPCs, and Model Selection
## 2.7 Discussion

